---
title: "R Tutorial"
output: html_notebook
---

This is Amos's R-Tutorial with all kinds of links and tricks accumulated over the last few years

to stop a script running - press Esc key or press the red stop sign on the console top line.

To clear all variables from previous runs:

```{r}
rm(list=ls()) # comes from remove

# to clear specific variables, but not all
rm(y, time1, time2, group, subject)

```
# In RStudio:
type ALT- to get the <-  symbol in one stroke
(Although the = assignment works too, other R users are not used to it, so avoid)

 Run Selection Ctr-Enter
 Run from here till end Ctr-Alt-E
 Run from begining till here Ctr-Alt-E
 
 to debug a function, enter browser()  where you want it to stop, then you can query everything
 
If you get duplicating commands on the Console, restart the session. SHould take care of problem. may cause problems when running certain scripts.


```{r}
```

# Useful links
 R Markdown Notebook cheat sheet:
 https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

 general brief tutorial, also with many functions 
 http://www.endmemo.com/program/R/index.php

 best tutorial on loops and vectorization
 https://www.datacamp.com/community/tutorials/tutorial-on-loops-in-r

 explains vectorized if-statement
 https://www.r-bloggers.com/for-loops-and-how-to-avoid-them/

 explains the apply function to several columns in a data set at once
 http://rforpublichealth.blogspot.co.il/2012/09/the-infamous-apply-function.html
 https://www.r-bloggers.com/apply-lapply-rapply-sapply-functions-in-r/

 Tutorial on Flow control and other R commands - very clear with e.g.'s
 https://www.programiz.com/r-programming/if-else-statement
 Tutorial on functions
 https://www.datacamp.com/community/tutorials/functions-in-r-a-tutorial

 a short page on data types:
 https://www.tutorialspoint.com/r/r_data_types.htm

 a long, but very clear page on data types:
 http://petewerner.blogspot.com/2013/09/type-conversion-and-you-or-and-r.html

 best tutorial for data frames creation, manipulation and data type conversion
 https://www.datacamp.com/community/tutorials/15-easy-solutions-data-frame-problems-r

 Good example for using data.table
 https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html 
 
 Tutorial about lists - elements of different types 
 https://www.tutorialspoint.com/r/r_lists.htm

 List useful commands for cummunity analysis, diversity indices and their data management
 http://ichthyology.usm.edu/courses/multivariate/data_management.R

 list of all other tutorials of datacamp, including importing from excel, other import data, plots, and more
 https://www.datacamp.com/community/tutorials

 General Tutorial for R. See advanced level below
 https://www.r-bloggers.com/how-to-learn-r-2/

 companion website for ?Advanced R?, Chapman & Hall?s book, designed to improve programming skills
 http://adv-r.had.co.nz/

 Cheat sheet for dplyr and tidyr packages - Data wrangling, combine dataframes etc
 http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
 Best Tutorial for dplyr:
 http://genomicsclass.github.io/book/pages/dplyr_tutorial.html
 tutorial for both dplyr and tidyr:
 https://rpubs.com/bradleyboehmke/data_wrangling

 for built-in functions see:
 http://www.statmethods.net/management/functions.html

 to work with dates see:
http://www.statmethods.net/input/dates.html

# ############### Good tutorials for plots  #######################
 http://www.cookbook-r.com/Graphs/

 Tutorial for R base graphics:
 http://rstudio-pubs-static.s3.amazonaws.com/7953_4e3efd5b9415444ca065b1167862c349.html

 A gentle introduction to ggplot
 https://rpubs.com/karagawa/ggplot2

 Cheatsheet to ggplot2
 https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf


 Beautiful plotting in R: A ggplot2 cheatsheet
 http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/ 

 general ggplot customization
 http://www.sthda.com/english/articles/32-r-graphics-essentials/125-ggplot-cheat-sheet-for-great-customization/#at_pco=smlre-1.0&at_si=5a8ec70c6213e72a&at_ab=per-2&at_pos=1&at_tot=4


 The following is a general catalog for many plotting needs:
 http://www.sthda.com/english/wiki/ggplot2-essentials

 Remember that ggplot2 requires data in long format. See far below in tutorial for converting wide to long

 A few useful pages from that catalog by www.sthda.com
 http://www.sthda.com/english/articles/32-r-graphics-essentials/132-plot-grouped-data-box-plot-bar-plot-and-more/



 http://www.sthda.com/english/articles/32-r-graphics-essentials/133-plot-one-variable-frequency-graph-density-distribution-and-more/

 http://www.sthda.com/english/articles/32-r-graphics-essentials/125-ggplot-cheat-sheet-for-great-customization/


 Very detailed and with many examples for plots
 http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization#at_pco=smlre-1.0&at_si=59995eafbd121a8b&at_ab=per-2&at_pos=3&at_tot=4


 Bar Plots
 http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization#barplot-with-error-bars

 Box plots
 http://www.sthda.com/english/wiki/ggplot2-boxplot-easy-box-and-whisker-plots-maker-function?url#at_pco=smlre-1.0&at_si=5a8ec70c6213e72a&at_ab=per-2&at_pos=2&at_tot=4


 some interesting plots ready for publication - histograms in certain orders etc
 http://www.sthda.com/english/rpkgs/ggpubr/


 Very detailed for multiple plots in one page
 http://www.sthda.com/english/wiki/ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page


 Photos of irises from the data set "iris"
 http://statlab.uni-heidelberg.de/data/iris/
 Deatils of the most common built-in data sets in R
 http://www.sthda.com/english/wiki/r-built-in-data-sets

 A bit of assistance in formatting R Notebooks
 https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

# ############## STATISTICS #######################

 an excellent Stats book - with lot's of simple methods including non-parametric
 https://rcompanion.org/rcompanion/
 Has also a pdf version of the book (may be in a different order of the chapters)
 https://rcompanion.org/documents/RCompanionBioStatistics.pdf
 
 2 examples of useful staff:
 
 Non-parametric regression
 https://rcompanion.org/handbook/F_12.html

 Confidence Interval calculation - parametric or by bootstrap - very clear
 http://rcompanion.org/handbook/C_03.html

 T-test, testing assumptions and various versions
 http://www.sthda.com/english/wiki/t-test-analysis-is-it-always-correct-to-compare-means#how-to-test-the-normality-of-data

 Randomization test of paired samples
 https://www.uvm.edu/~dhowell/StatPages/ResamplingWithR/RandomMatchedSample/RandomMatchedSampleR.html
 
 Randomization test for independent samples
 https://www.uvm.edu/~dhowell/StatPages/R/RandomizationTestsWithR/Random2Sample/TwoIndependentSamplesR.html

 correlation plots with visualization
  http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2

 http://www.sthda.com/english/wiki/correlation-matrix-an-r-function-to-do-all-you-need#at_pco=smlre-1.0&at_si=5a8ec70c6213e72a&at_ab=per-2&at_pos=3&at_tot=4
 
 https://www.r-bloggers.com/r-tutorial-series-regression-with-interaction-variables/  Demonstrates how to use the interactions term

 Visualizing Likert data (in questionnairs - e.g., 5 answers: strongly agree, agree, etc)
 For Analysis - see https://www.joe.org/joe/2012april/tt2.php (the idea - no R code)
 Also - https://pdfs.semanticscholar.org/e2ad/36059a7894c89ea566811eb82ad92acd88a7.pdf (the same as previous - no code)
 https://rcompanion.org/handbook/E_03.html

 Paired Wilcoxon analysis and plots
 http://www.sthda.com/english/wiki/paired-samples-wilcoxon-test-in-r

 ANOVA's, Repeated Measures, Testing assumptions for these + Non-parametrics
 EZ package
 https://cran.r-project.org/web/packages/ez/ez.pdf
 to remember fixed vs. random effects: https://www.researchgate.net/post/What_is_the_difference_between_fixed_effect_and_random_effect_in_repeated_measures_ANOVA

 https://www.r-bloggers.com/r-tutorial-series-two-way-anova-with-unequal-sample-sizes/
 tutorial about ANOVA with unequal sample sizes

 permutation tests for linear models - ANOVA, ANCOVA, Regression. One example - only ANCOVA, but clear
https://statmethods.wordpress.com/2012/05/21/permutation-tests-in-r/

 Log-linear model - see Shaked Notebook on mounds
 https://data.library.virginia.edu/an-introduction-to-loglinear-models/
 
 The output is not explained in this tutorial, so can use the following for interpreting the summary + very nice visualization
 https://www.dataquest.io/blog/tutorial-poisson-regression-in-r/
 
 But the most comprehensive for all categorical analyses with many options incl. one type of visualization is here:
 https://r-forge.r-project.org/scm/viewvc.php/*checkout*/pkg/inst/doc/vcd-tutorial.pdf?revision=54&root=vcdextra&pathrev=54

 survival plots (think chameleons)
 http://www.sthda.com/english/wiki/survminer-r-package-survival-data-analysis-and-visualization

 PCA, CA and others
  http://www.sthda.com/english/wiki/factoextra-r-package-easy-multivariate-data-analyses-and-elegant-visualization

# ########### Built-in data sets in R or in specific packages ########

 the common iris and mtcars - are part of R.

```{r}
data("iris")  # use the built-in data set, or the one in the package you are using 
str(iris)   # see the structure of the data set
iris  # have a look at the data set

```

# ################## Packages ##############

Note than in R " and ' are compatible (but don't mix together). Single and double quotes delimit character constants. They can be used interchangeably but double quotes are preferred.


```{r}
# In case you are not sure if package is installed - this will only install it if missing
if (!require(xlsx)) install.packages('xlsx') # use exact syntax as in the example, including teh quote marks!

# then you need to call it as usual
library("xlsx")
library(xlsx) # should work also without the quote marks

detach(package:xlsx)  # to remove a package you called in a specific session (sometimes there are incompatibilities, or if you want to know if it is really needed)
detach('package:xlsx', unload = TRUE, character.only = TRUE) # another version suggested to fully remove remains
```


Working Directory issues (less relevant for notebook users)
```{r}
getwd() # to check which is the current working directory (unless you use a notebook, and then the directory is always the directory where the notebook is)

# to change to another working directory (but read above - can't change in a notebook the working directory:
setwd("U:\\Backup 15 Nov 11\\Amos - 21 Apr 2010\\Ornatus advance\\Aegyptius Einat\\R analysis Garden")
 # make sure you have the double \\

```

# ################ Input #################
Different Input methods

Option 1a - find data file yourself
```{r}
 
dat<-read.csv(file.choose())

```

Option 1b - read an excel file. See more options in: 
http://www.sthda.com/english/wiki/reading-data-from-excel-files-xls-xlsx-into-r#importing-excel-files-using-xlsx-package
for this you need to use the package xlsx (that was installed in previous chunk)
In general, there are sometimes unclear problems with data imported directly from Excel... Best is to import from .csv
But if you want directly from Excel - this is the way

```{r}
my_data <- read.xlsx(file.choose(), 1)  # read first sheet (an advantage you don't have in read.csv)
# alternative, for large data sets:
my_data <- read.xlsx2(file.choose(), 1)  # Different command, still in the same xlsx package

```

option 1c - use Import Dataset in Top Right window in RStudio. csv or excel.

Option 2a - read a known .csv file

```{r}
dat<-read.csv("chamData4.csv")  # a copy of this file must be in the working directory, or you need to specify the full path, as below (even in a Notebook)

dat<-read.csv("U:\\from Think Pad\\Documents from Think Pad\\R\\My scripts from all over\\ScriptsCatalog.csv")  # or you need to specify the full path. Remember the double \\ or use one in opposite direction /

dat<-read.csv("U:/from Think Pad/Documents from Think Pad/R/My scripts from all over/ScriptsCatalog.csv")  #  or use only one in opposite direction /

```


Option 2b - read a known Excel file

Be aware that it comes as a data frame, and sometimes it causes problems, as it does not recognize a column as numerical vector for some reason. Does not happen when you read as .csv

This example with a different package, unlike previous
```{r}
library(readxl)  # enables to import Excel files directly without .csv
Trees <- read_excel("Trees.xlsx") # must be in the working directory

```
Option 3 - read from clipboard

does not work in Mac, or you need to write it differently

```{r}
dat<-read.delim("clipboard")
```

Option 4 - enter data within the script

```{r}
NebkaCl <- c(41  ,  300,  250,	210	,268,	338)
NebkaOp <- c(325  ,145	,55,	277,	197,	285	,328	,276	,312	,290)
```


replace missing values (called NA in R) to 0 (vectorized)

replace negative values to something else

```{r}
dat[is.na(d)] <- 0
dat[dat<0] <- 0

```
Continue from line 281 in original file

```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


# Note that <- is equivalent in most cases to = (preferred by advanced programmers, so if you want to consult with them...). Note that sometimes you may want to use the symbol <<- This will make the variable global, which might be important if you want to read it from within a function, or from loops.

dev.off() # to close all plots created so far in a session

?print # to get help on a function name. The help will appear in lower right window


library(Hmisc)
rcorr(as.matrix(Merged[2:4]), type="spearman") # type can be pearson or spearman
# Alternatively, 
# See in Gil's Soc Network for a very nice visualization:
library("PerformanceAnalytics")
my_data <- Merged[2:4]
chart.Correlation(my_data, histogram=TRUE, method = "spearman", pch=10)



```
